# Analysis of Longitudinal Data

## Chapter 8: Graphical Displays and summary measure approach

### The data=

The data set that will be used is from a nutrition study in three groups of rats (Crowder and Hand 1990). The weight of each rat was measured repeatedly 11 times every 7 weeks. The main objective is to determine if the different diets have effect on rats growth measured as weight.

The **rats** data set can be access using this [link](https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt). The data set have 16 observations and 13 columns. It has a wide format wherein each column contains one repeated measurement for the same rat.

```{r}
RATS<-read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt",
                 sep = "", header = TRUE)

dim(RATS); str(RATS); summary(RATS)
```

Notice after loading the data set, R did not recognized the first two columns (**$ID** and **$Group**) as categorical variables. It has to be manually changed into categorical variables (**as.factor**). Furthermore, the data has to be converted into a long format by collapsing the repeated measurements into one column **weight**.

```{r message = FALSE}
library(dplyr)
library(tidyr)

lRATS <- RATS %>%
  mutate(
    fID = as.factor(ID),                        # create a new column $fID, which contains categorical variable
    fGroup = as.factor(Group)                   # create a new column $fGroup which contains categorical variable 
  ) %>% 
  dplyr::select(c(-ID, -Group)) %>%                          # Remove the columns ID and Group
  gather(key = time, value = weight, -fID, -fGroup) %>%      # Reformat the data set from wide to long
  dplyr::mutate(week = as.integer(substr(time, 3,4)))        # create a new column $week which contains the numerical part of the column $time

#' lRATS <- long format of RATS data set
str(lRATS)

```

### Graphical overview

We can further explore the data by creating plots to discover patterns. The first plot shows the weight of each rats overtime. The plot shows that there is uneven number of rats on each diet treatment. Furthermore, the initial weight of the rats vary across treatment. Overall, all the rats gained weight regardless of what type of diet they received. 


```{r message = FALSE}
library(ggplot2)

p1 <- ggplot(lRATS, aes(x = week, y = weight)) + geom_line(aes(color = fGroup))
p1 <- p1 + facet_wrap( ~ fID, scale = 'free')
p1 <- p1 + labs(color = 'Diet', caption = 'Fig 1. Individual rat weight measurements over  time.')
p1 <- p1 + theme(plot.caption = element_text(hjust = 0, size = 14))
                 
p1

```


The boxplot shows the average weight gained by the rats in Diet 2 is not as significant compared to the other two diets (**Fig 2**). Furthermore, extreme weight values are persistent in all treatment over time.


```{r}
#' Boxplot
p2 <- ggplot(lRATS, aes(x = week, y = weight, group = week)) + geom_boxplot(aes(color = fGroup))
p2 <- p2 + labs(color = 'Diet', caption = 'Fig 2. Boxplot for the RATS data. Each panel is for every diet treatment')
p2 <- p2 + facet_grid( fGroup ~ ., scale = 'free') 
p2 <- p2 + theme(plot.caption = element_text(hjust = 0, size = 14))
p2
                 
```


The difference in weight gained among the different diet is more pronounce in **Fig 3**. In addition, variation is high across treatment but weight gain is steeper in diet 1.

```{r message = FALSE}
#' Summary
library(Rmisc)
aveRats <- summarySE(data = lRATS, measurevar = 'weight', groupvars = c('week', 'fGroup'))

p3 <- ggplot(aveRats, aes(x = week, y = weight)) + geom_point(aes(shape = fGroup))
p3 <- p3 + geom_ribbon(aes(ymin = weight - ci, ymax = weight + ci), fill = "grey70", alpha = 0.5) + geom_line(aes(color = fGroup))
p3 <- p3 + geom_errorbar(aes(ymin = weight - se, ymax = weight + se))
p3 <- p3 + theme(legend.position = 'none',
                 plot.caption = element_text(hjust = 0, size = 14)) +
  labs(caption = 'Fig 3. Average weight per week for each diet treatment with confidence interval. error bars = Standard Error')
p3 <- p3 + facet_wrap(fGroup ~ ., scale = 'free')
p3
```


**Fig 4** shows boxplots of rat's average weight from week 1 - 64 for each treatment.Rats that receives diet 1 is smaller compared to the rats which received the other treatment. Furthermore, rat weight in diet 2 are skewed towards smaller rats. 


```{r}
#' average weight over week 0 - 64
p4 <- ggplot(lRATS, aes(x = fGroup, y = weight, group = fGroup)) + geom_boxplot(aes(color = fGroup))
p4 <- p4 + labs(x = 'Diet', y = 'weight (week 1 -64)', caption = 'Fig 4. Average weight for each treatment over 64 weeks')
p4 <- p4 + theme(axis.title = element_text(size = 14),
                 legend.position = 'none', plot.caption = element_text(hjust = 0, size = 14))
p4
```


### Fitting a model

Analysis of variance suggests that the weight of the rats from each diet are significantly different. Post hoc test suggest the pairwise difference among the diet tretment are all significant as well.
```{r message = FALSE}
library(pander)
#' anova
fit1 <- aov(weight ~ fGroup, data = lRATS)
pander(summary(fit1))
plot(TukeyHSD(fit1))
```


A linear model using weight as response and diet as fixed affect was fitted.The results supports the result from ANOVA that diet linearly affects the rat's weight (**R2 = 0.92**).Diet 3 seems to increase the weight of the rat the most. The **R2** ar 

```{r message = FALSE}
#' linear model
fit2 <- lm(weight ~ fGroup, data = lRATS)
pander(summary(fit2))

```

## Chapter 9: Linear mixed effects models

### The data

The **BPRS** data will be used on this exercise (Davis 2002). The data set contains information about 40 males randomly assigned to one of the two treatments. Initial psychiatric rating scale (BPRS) was measured before the treatment (Week 0), and measured thereafter at weekly interval for 8 weeks. The scale is used to evaluate if a person has schizophrenia. 

The scatterplots clearly shows the repeated measure nature of the BPRS scores for both treatment.

```{r message = FALSE}
#' upload data
BPRS<-read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt",
                 sep = "", header = TRUE)

#' explore data
dim(BPRS); str(BPRS); summary(BPRS)
head(BPRS)

#' scatterplot
#BPRS$subject<-as.factor(BPRS$subject)
library(GGally)
ggpairs(BPRS[,3:11], ggplot2::aes(color = as.factor(BPRS$treatment)))

```

### Converting wide to long data

The data set is in wide format and needs to change to long format before further analyses.

```{r message = FALSE}
library(dplyr)
library(tidyr)
lBPRS <- BPRS %>%
  mutate(
    ftreatment = as.factor(treatment),
    fsubject = as.factor(subject)
  ) %>% 
  dplyr::select(c(-treatment, -subject)) %>%
  gather(key = weeks, value = bprs, -ftreatment, -fsubject) %>%
  dplyr::mutate(week = as.integer(substr(weeks, 5,5)))

#' explore data
head(lBPRS, 5)
str(lBPRS)
lBPRS$ID <- rep(1:40, 9) %>% as.factor()
```

### Graphical overview

Plotting the bprs scores over time for each subject shows decreasing trend in most cases for both treatment.
```{r message = FALSE}
#' Graphical overview
p1 <- ggplot(lBPRS, aes(x = week, y = bprs, group = ID)) + geom_line(aes(color = ftreatment))
p1 <- p1 + labs(color = 'Treatment', caption = 'Fig 1. Individual BPRS profile by treatment group')
p1 <- p1 + theme(plot.caption = element_text(hjust = 0, size = 14)) +
  facet_wrap(~ ftreatment, scale = 'free') 
p1

```


### Linear regression model

Different models were tested to predict bprs scores. By comparing the AIC scores of the three models, the best model was found to be with random intercept and random slope without interaction between the fixed variables. Furthermore, the treatment did not result to significant difference in bprs scores.

```{r message = FALSE}
#'linear model
library(lme4)
library(sjmisc)
library(sjPlot)
library(pander)
fit1 <- lm(bprs ~ week + ftreatment, data = lBPRS)
pander(summary(fit1))

#' subject as random effect
fit2 <- lmer(bprs ~ week + ftreatment + (1 | ID), data = lBPRS, REML = FALSE)


fit3 <- lmer(bprs ~ week + ftreatment + (week | ID), data = lBPRS, REML = FALSE)


fit4 <- lmer(bprs ~ week * ftreatment + (week | ID), data = lBPRS, REML = FALSE)
tab_model(fit2, fit3, fit4,
          dv.labels = c('Random Intercept', 'Random Intercept & Slope', 'Randrom Intercept & Slope + Interaction'))

pander(anova(fit2, fit3, fit4))

```

## Predict

BPRS scores can be predicted using the selected model. The results from **t.test** between the predicted value to the actual bprs scores suggest the model can fairly predict BPRS scores (**Fig 2a and 2b**).


```{r}
lBPRS$pred <- predict(fit3, type = 'response') %>% as.integer()
str(lBPRS)

t.test(lBPRS$pred, lBPRS$bprs)

llBPRS <- lBPRS %>% gather(key = type, value = bprs, -ftreatment, -fsubject, -weeks, -week, -ID)

llBPRS <- llBPRS %>% mutate(
  type = as.factor(type),
  week = as.factor(week),
  ID = as.factor(ID)
)
str(llBPRS)

p2a <- ggplot(subset(llBPRS, ftreatment == '1'), aes(x = week, y = bprs, group = ID)) + geom_line(aes(color = type))
p2a <- p2a + labs(caption = 'Fig 2a. Observed vs Predicted BPRS points for treatment 1') +
  facet_grid( type ~ fsubject) +
  theme(plot.caption = element_text(hjust = 0, size = 14))
p2a

p2b <- ggplot(subset(llBPRS, ftreatment == '2'), aes(x = week, y = bprs, group = ID)) + geom_line(aes(color = type))
p2b <- p2b + labs(caption = 'Fig 2b. Observed vs Predicted BPRS points for treatment 2') +
  facet_grid( type ~ fsubject) +
  theme(plot.caption = element_text(hjust = 0, size = 14))
p2b


```

