# Clustering and Classification

## Data exploration

The **Boston** data set from MASS package will be used for this exercise. The data is stored on this [archive](http://lib.stat.cmu.edu/datasets/boston) and more information is available on this [link](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

The data was originally used in this [paper](https://www.sciencedirect.com/science/article/pii/0095069678900062). It contains information about housing in Boston, Massachusetts collected by U.S. Census Service.


```{r}
library(MASS)

data("Boston")

#' Explore data structure
#' it has 506 observations and 14 attributes/variables
dim(Boston)
str(Boston)

#'check for nas
colSums(is.na(Boston))

#' check summaries of each variables
summary(Boston)

pairs(Boston)
```

## Standardizing the data set



```{r}
library(dplyr)
#' Scale the original data set
#' this is results to each attibute to have mean = 0
scaled.B<- scale(Boston) %>% as.data.frame()

#'
summary(scaled.B)

#'
class(scaled.B)
```

## Creating categorical variable **crime**

```{r}
#' print summary of variable 'crim'
summary(scaled.B$crim)

#' create quantile vector of crim
bins<- quantile(scaled.B$crim)

#' create  categorical variable 'crime'
crime <- cut(scaled.B$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
scaled.B <- dplyr::select(scaled.B, -crim) %>% data.frame(scaled.B, crime)

colnames(scaled.B)
head(scaled.B, 5)


```

## Creating **training** and **test** subset


```{r}
# number of rows in the Boston dataset 
n <- nrow(scaled.B)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- scaled.B[ind,]

#nrow(train) == n*.8

# create test set 
test <- scaled.B[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

```

## Fitting **linear discriminat analysis**

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ . , data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

```


## Predict the classes
```{r}

# save the correct classes from test data
# correct_classes <- test$crime

# remove the crime variable from test data
# test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```

## Clustering

```{r}
Boston.std<- scale(Boston) %>% as.data.frame()

# euclidean distance matrix
dist_eu <- dist(Boston.std)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(Boston, method = "manhattan")

# look at the summary of the distances
summary(dist_man)

# k-means clustering
km <-kmeans(Boston.std, centers = 3)

# plot the Boston dataset with clusters
pairs(Boston.std, col = km$cluster)



```




